Comment on Cumulative Excess Deaths in New Zealand in the COVID-19 Era: Biases from Ignoring Changes in Population Growth Rates

### 2024-07-12 rawfiles/base_content.qmd
tidying repo to match final version also added Excel example

---
date: last-modified
title: "Comment on Cumulative Excess Deaths in New Zealand in the COVID-19 Era: Biases from Ignoring Changes in Population Growth Rates"
author: "David Hood"
execute:
  keep-md: true
format:
  pdf:
    fig-pos: 'h'
bibliography: references.bib
csl: apa.csl
---

This is the Github repo for my letter of comment on Gibson's article. The letter references the code data files in the support folder available here. For context, the final draft/ preprint of the letter is included below, though this final draft may differ slightly from any published version.

------------------------------------------------------------------------

For hundreds of years, correctly measuring mortality has been a concern of actuaries, demographers, and health researchers. Accurate measurement should form the basis of insurance risk pricing and government policy. Writing in this journal, Gibson [@gibson2024] asserts that New Zealand authorities' measurement of mortality in the pandemic period is based on a flawed metric. Gibson's own metric, based on regression of deaths with an additive component of total population, is asserted as a more accurate model and produces approximately 4% excess mortality for the 2020-2022 period. If New Zealand authorities are using flawed metrics with significantly different results to reality, this would have profound governance and policy implications.

In particular, the paper asserts two broad claims:

@@

For centuries, it has been established professional practice to measure mortality using age-adjusted death rates where such data is available [@price1773]. This method supersedes the use of Crude Mortality Rates[^2] by acknowledging that different ages have different risks of death, so comparing age-based risk of death on a uniform basis. In New Zealand, tutorials on making mortality comparisons by using death rates by age and standardising to a reference population have been published since the 1890s [@didsbury1891].

@@

While calculation of New Zealand's cumulative excess mortality for 2020-2022 can vary by a couple of percentage points depending on the exact method adopted, using an age-standardised death rates method the UK's Institute and Faculty of Actuaries finds New Zealand's excess mortality from week 10 2020 to the end of 2022 to be -4% (that there are 4% fewer deaths than would be expected based on pre-COVID-19 trends) [@instituteandfacultyofactuaries2024].

The Our World in Data graph, and the Karlinsky and Kobak method that underlies it, is an approximation (if it is a better or worse approximation to Gibson's method will be discussed later). While recognising the Our World In Data graph is an approximation, there are still several practically useful implicit features about it:

@@

The existence of centuries of mortality measurement establishes that it is best practice in mortality calculations to recognise that people of different ages die at different rates not mentioned in Gibson's article. While New Zealand medical professionals have used the Karlinsky and Kobak method/ Our World in Data graphs in media presentations, it is not established if that was for the designed practical purposes while still being aware of more accurate methods or because K & K was considered definitive. We can check if New Zealand medical professionals are aware of best practice methods by asking them.

@@

Though not mentioned in Gibson's paper, the commonest method for measuring deaths in relation to total population (without aging) is the Crude Mortality Rate (deaths among the total resident population / total living resident population). This is available for any country producing both an annual count of resident deaths and an annual resident population count. A 2015-2019 linear regression of the annual CMR[^8] gives an 2020-2022 cumulative excess of 0%.

@@

Gibson's decision not to apply total population but not age to the model minimises an expected increase in deaths of an approximate order of magnitude greater than the effects of migration. But all these complexities are avoided by incorporating both migration and aging by using age standardised mortality.

Gibson asserted that it was not important to incorporate aging "If societal aging was a cause of the increased number of deaths, as Gabel and Knox (2023) put forward as a reason for the excess mortality in 2022, it should also show up prior to the COVID-19 era, given that societal aging is a long-term process." (Gibson, 2024, pp.5)

The assumption that aging is a long-term process so should show up as a long slow process only applies if there are no sudden demographic shifts among age groups that alter the risk of death. As the pre-COVID-19 era, tautologically, the baseline for COVID-19 comparisons and trends fitted to it automatically best fit the data of the period, to see the pre-COVID-19 rise requires using a slightly earlier baseline.

The number of 75 and older residents has been growing at an accelerating rate in the decade prior to COVID-19 (see Figure 2). This matches a "sudden" demographic shift of an acceleration in the age distribution that began decades earlier and naturally aged into ranges that play a more significant role in total deaths.

@@

But the dramatic increase in elderly is only one of several interacting factors contributing to mortality. Others are factors influencing the age specific risk environment. A more direct measure of the importance of age stratification is the difference in outcome between the Crude Mortality Rate (ignoring age structure) and the Age Standardised Mortality Rate (based on age stratified death rates). The difference in excess mortality between those two methods of 2% to 4% is exactly, and solely, the difference caused by not using age specific death rates.

Of more surprise than the comparison to age standardised methods is that Gibson's results for 2020-22 less well approximate the crude mortality rate than Karlinsky & Kobak, since the crude mortality rate is the normal method of comparing death rates while ignoring the effects of age. While there is not enough information in the paper to exactly replicate Gibson's method and achieve the same outcome, part of the reason for the difference seems likely to be the unusual step of treating population as an additive contributor to deaths in his modification of Karlinsky & Kobak. But, compared to the differences with methods acknowledging age, such problems are minor.

@@

From comparison to best practice ASMR methods, K&K are closer in result than Gibson's method. Further, given the results of ASMR, New Zealand and international professionals have a sound basis for determining New Zealand's excess mortality as negative for the covid period.

>>>>

---
bibliography: references.bib
csl: apa.csl
---

For hundreds of years, correctly measuring mortality has been a concern of actuaries, demographers, and health researchers. Accurate measurement should form the basis of insurance risk pricing and government policy. Writing in this journal, Gibson [@gibson2024] asserts that New Zealand authorities’ measurement of mortality in the pandemic period is based on a flawed metric. Gibson’s own metric, based on regression of deaths with an additive component of total population, is asserted as a more accurate model and produces approximately 4% excess mortality for the 2020-2022 period. If New Zealand authorities are using flawed metrics with significantly different results to reality, the thousands of unnoticed deaths would have profound governance and policy implications.

In particular, I interpret the paper as asserting two broad claims:

@@

It has been established professional practice for centuries to measure mortality using age-adjusted death rates where such data is available [@price1773]. This method supersedes the use of Crude Mortality Rates[^2] by acknowledging that different ages have different risks of death, so comparing age-based risk of death on a uniform basis. In New Zealand, tutorials on making mortality comparisons by using death rates by age and standardising to a reference population have been published since the 1890s [@didsbury1891].

@@

While calculation of New Zealand’s cumulative excess mortality for 2020-2022 can vary by a couple of percentage points depending on the exact method adopted, using an age-standardised death rates method the UK’s Institute and Faculty of Actuaries finds New Zealand’s excess mortality from week 10 2020 to the end of 2022 to be -4% (that there are 4% fewer deaths than would be expected based on pre-COVID-19 trends) [@instituteandfacultyofactuaries2024].

The Our World in Data graph, and the Karlinsky and Kobak method that underlies it, is an approximation (if it is a better or worse approximation to Gibson’s method will be discussed later). While recognising the Our World In Data graph is an approximation, there are still several practically useful implicit features about it:

@@

Centuries of mortality measurement establishes that it is best practice in mortality calculations to recognise that people of different ages die at different rates. While New Zealand medical professionals have used the Karlinsky and Kobak method/ Our World in Data graphs in media presentations, it is not established if that was for the above practical benefits while still being aware of more accurate methods, or because K & K was considered definitive. We can check if New Zealand medical professionals are aware of best practice methods by asking them.

@@

Though not mentioned in Gibson's paper, the commonest method for measuring deaths in relation to total population (without adjusting for aging) is the Crude Mortality Rate (deaths among the total resident population / total living resident population). This is available for any country producing both an annual count of resident deaths and an annual resident population count. A 2015-2019 linear regression of the annual CMR[^8] gives an 2020-2022 cumulative excess of 0%. This establishes that a method not adjusting for age, as might be logically expected in a country with an aging population, gives a higher value for excess than a method that adjusts for age.

@@

Gibson’s decision not to apply total population but not age to the model minimises an expected increase in deaths of an approximate order of magnitude greater than the effects of migration. But all these complexities are avoided by incorporating both migration and aging by using age standardised mortality.

Gibson asserted that it was not important to incorporate aging “If societal aging was a cause of the increased number of deaths, as Gabel and Knox (2023) put forward as a reason for the excess mortality in 2022, it should also show up prior to the COVID-19 era, given that societal aging is a long-term process.” (Gibson, 2024, pp.5)

The assumption that aging is a long-term process so should show up as a long slow process only applies if there are no sudden demographic shifts among age groups that alter the risk of death. In the pre-COVID-19 era, tautologically, the baseline for COVID-19 comparisons and trends are automatically best fit the data of the pre-covid period, to see the pre-COVID-19 rise is aided by using a slightly earlier baseline.

The number of 75 and older residents has been growing at an accelerating rate in the decade prior to COVID-19 (see Figure 2). This matches a “sudden” demographic shift of an acceleration in the age distribution that began decades earlier and naturally aged into ranges that play a more significant role in total deaths.

@@

But the dramatic increase in elderly is only one of several interacting factors contributing to mortality. Others are factors influencing the age specific risk environment. A more direct measure of the importance of age stratification is the difference in outcome between the Crude Mortality Rate (ignoring age structure) and the Age Standardised Mortality Rate (based on age stratified death rates). The difference in excess mortality between those two methods of 2% to 4% is the difference caused by not using age specific death rates.

Of more surprise than the comparison to age standardised methods is that Gibson’s results for 2020-22 less well approximate the crude mortality rate than Karlinsky & Kobak, since the crude mortality rate is the normal method of comparing death rates while ignoring the effects of age. This may be the result of the unusual step of treating population as an additive contributor to deaths in his modification of Karlinsky & Kobak. But, compared to the differences with methods acknowledging age, such problems are minor.

@@

From comparison to best practice ASMR methods, K&K are closer in result than Gibson’s method. Further, given the results of ASMR, New Zealand and international professionals have a sound basis for determining New Zealand’s excess mortality as negative for the covid period.
